---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rfars

<!-- badges: start -->

<!-- badges: end -->

The goal of rfars is to simplify the process of analyzing FARS data. The
[Fatality and Injury Reporting System Tool](https://cdan.dot.gov/query) allows
users to generate queries, and can produce simple tables and graphs. This
suffices for simple analysis, but often leaves researchers wanting more. Digging
any deeper, however, involves a time-consuming process of downloading annual ZIP
files and attempting to stitch them together - after combing through the immense
data dictionary to determine the required variables and table names. rfars
allows users to download five years of FARS data with just two lines of code.
The result is a full, rich dataset ready for mapping, modeling, and other
downstream analysis. Helper functions are also provided to produce common counts
and comparisons.

## Installation

You can install the latest version of rfars from [GitHub](https://github.com/)
with:

``` r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

## Getting Data

Here we import 5 years of data for Virginia. Note that get_fars requires
your permission to download the ZIP files from NHTSA and save the prepared files
on your hard drive. This only has to be run once and defaults to saving everything
in the current working directory. The use_fars function looks in that directory
for certain files and compiles them into one. 

```{r example}
library(rfars)
library(dplyr)

# get_fars(years = 2016:2020, states="Virginia")

myFARS <- use_fars()

#head(myFARS)
```

You can review the list of variables to help guide your analysis with:

``` r
View(fars_varnames)
```

## Counts

```{r, fig.height=2.2}
library(ggplot2)

counts(myFARS,
       what = "crashes",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Crashes in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Rural Fatalities in Virginia")
```

Let's compare the number of speeding-related fatalities in rural and urban places.

```{r, fig.height=2.4}
bind_rows(
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Rural"),
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "urban",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Urban")
    ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~where) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

## Mapping

```{r}

```

## Modeling

```{r}

```


