---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi = 500
)
```

# rfars

<!-- badges: start -->

<!-- badges: end -->

The goal of rfars is to simplify the process of analyzing FARS data. The
[Fatality and Injury Reporting System Tool](https://cdan.dot.gov/query) allows
users to generate queries, and can produce simple tables and graphs. This
suffices for simple analysis, but often leaves researchers wanting more. Digging
any deeper, however, involves a time-consuming process of downloading annual ZIP
files and attempting to stitch them together - after first combing through the
immense [data
dictionary](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813254)
to determine the required variables and table names. `rfars`allows users to
download five years of FARS data with just two lines of code. The result is a
full, rich dataset ready for mapping, modeling, and other downstream analysis.
Helper functions are also provided to produce common counts and comparisons. A
companion package `rfarsplus`provides exposure data and facilitates the
calculation of various rates.

## Installation

You can install the latest version of `rfars`from [GitHub](https://github.com/)
with:

``` r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

```{r}
library(rfars)
library(dplyr)
library(ggplot2)
```


## Getting and Using FARS Data

Use the `get_fars()` function to download the ZIP files from NHTSA and save 
the prepared files to your hard drive. This only has to be run once and defaults 
to saving everything in the current working directory. Below we import 5 years 
of data for Virginia. Note that `get_fars()` requires your permission to 
download the ZIP files and store prepared CSVs locally. 

```{r, eval=FALSE}
get_fars(years = 2016:2020, states="Virginia")
```

The `use_fars()` function looks in that directory for certain files and compiles 
them into a list of four data frames: `flat`, `multi_acc`, `multi_veh`, and 
`multi_per`. The `flat` file contains all variables for which there is just one
value per person, vehicle, or crash (e.g., age, travel speed, lighting). Each 
row in this table corresponds to a person involved in a crash. As there may be
multiple people and/or vehicles involved in one crash, some variable-values
are repeated within a crash or vehicle. Each crash is uniquely identified with
`id`, which is a combination of `year` and `st_case`. Note that `st_case` is not
unique across years, for example, `st_case` 510001 will appear in each year. The
`id` variable attempts to avoid this issue.

The `multi_` files contain those variables for which there may be multiple 
values for any entity (e.g., driver impairments, vehicle events, weather 
conditions at time of crash). Each table has the requisite data elements
corresponding to the entity: `multi_acc` includes `st_case` and `year`,
`multi_veh` adds `veh_no` (vehicle number), and `multi_per` adds `per_no` 
(person number). 

```{r}
myFARS <- use_fars("FARS data") 

str(myFARS)
```

You can review the list of variables to help guide your analysis with:

``` r
View(fars_varnames)
```

## Counts

A first step in many transportation analyses involves counting the number of
relevant crashes, fatalities, or people involved. `counts()` lets users specify a
time period and aggregation interval, and focus in on specific road users and
factors. It can be combined with `ggplot()` to quickly visualize counts.

```{r, fig.height=2.2}

counts(myFARS,
       what = "crashes",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Fatal Crashes in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Rural Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities\nin Rural Virginia")
```

We can combine two `counts()` results to make a comparison. Here we compare the 
number of speeding-related fatalities in rural and urban Virginia:

```{r, fig.height=2.4}
bind_rows(
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Rural"),
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "urban",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Urban")
    ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~where) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

## Mapping

We can take advantage of having access to the full data with maps. 

Here we map pedestrian and bicyclist fatalities in Virginia:

```{r}
library(leaflet)
library(leaflet.extras)

counts(
  myFARS, 
  what = "fatalities", 
  when = 2016:2020, 
  involved = "pedbike", 
  filterOnly = TRUE
  ) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 10) %>%
  addCircleMarkers(
    radius = 1.5,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```


Drug-related fatalities:

```{r}
counts(
  myFARS, 
  what = "fatalities", 
  when = 2016:2020, 
  involved = "drugs", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 10) %>%
  addCircleMarkers(
    radius = 1.5,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```


Older driver crashes:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  when = 2016:2020, 
  involved = "older driver", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 10) %>%
  addCircleMarkers(
    radius = 1.5,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```


Young drivers:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  when = 2016:2020, 
  involved = "young driver", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 10) %>%
  addCircleMarkers(
    radius = 1.5,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```




## Modeling

Having access to the full dataset also allows us to develop statistical models.
Here we fit a simple model of injury severity:

```{r}
#table(myFARS$flat$inj_sev)
#table(myFARS$flat$rest_mis)

myFARS$flat %>%
  mutate(kabco = case_when(inj_sev == "Fatal Injury (K)" ~ 4,
                           inj_sev %in% c("Suspected Serious Injury (A)", 
                                          "Suspected Serious Injury(A)") ~ 3,
                           inj_sev %in% c("Suspected Minor Injury (B)", 
                                          "Suspected Minor Injury(B)") ~ 2,
                           inj_sev == "Possible Injury (C)" ~ 1,
                           inj_sev == "No Apparent Injury (O)" ~ 0,
                           TRUE ~ as.numeric(NA)
                           ),
         age_n = gsub("\\D+","", age) %>% as.numeric()) %>%
  lm(kabco ~ age_n + rest_mis, data = .) %>%
  summary()
```
