---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%", fig.align = "center",
  dpi = 500
)
```

# rfars <img src="man/figures/logo.svg" align="right" alt="" width="120" />

<!-- badges: start -->
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/rfars)](https://cran.r-project.org/package=rfars)
[![R CMD Check](https://github.com/s87jackson/rfars/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/s87jackson/rfars/actions/workflows/R-CMD-check.yaml)
[![](https://cranlogs.r-pkg.org/badges/grand-total/rfars)](https://CRAN.R-project.org/package=rfars)
<!-- badges: end -->

The goal of `rfars` is to facilitate transportation safety analysis by
simplifying the process of extracting data from official crash databases. 
The [National Highway Traffic Safety Administration](https://www.nhtsa.gov/) collects and 
publishes a census of fatal crashes in the 
[Fatality Analysis Reporting System](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars) 
and a sample of fatal and non-fatal crashes in the
[Crash Report Sampling System](https://www.nhtsa.gov/crash-data-systems/crash-report-sampling-system) 
(an evolution of the [General Estimates System](https://www.nhtsa.gov/national-automotive-sampling-system/nass-general-estimates-system)).
The [Fatality and Injury Reporting System Tool](https://cdan.dot.gov/query) allows users to 
query these databases, and can produce simple tables and graphs. This suffices for
simple analysis, but often leaves researchers wanting more. Digging any deeper,
however, involves a time-consuming process of downloading annual ZIP files and
attempting to stitch them together - after first combing through immense data dictionaries 
to determine the required variables and table names. 

`rfars` allows users to download FARS and GES/CRSS data back to 2011 with just 
one line of code. The result is a full, rich dataset ready for mapping, modeling, 
and other downstream analysis. Helper functions are also provided to produce 
common counts and comparisons.

## Installation

You can install the latest version of `rfars` from [GitHub](https://github.com/)
with:

```r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

or the CRAN stable release with:

```r
install.packages("rfars")
```

Then load rfars and some helpful packages:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(rfars)
library(dplyr)
library(ggplot2)
library(leaflet)
library(leaflet.extras)
```

## Getting and Using FARS Data

Use the `get_fars()` function to bring FARS data into the current environment and optionally
save an RDS file for future use (faster than downloading anew). 
This is done by (a) downloading the data to a temporary directory, (b) downloading
to a permanent directory, or (c) importing from a permanent directory. After 
data is downloaded to a permanent directory, the function will look there rather
than downloading the data again. If a year of data is requested but not found,
R will ask your permission to download the missing data. 

Here we get three years of data for Virginia:

```{r}
myFARS <- get_fars(years = 2019:2021, states = "VA")
```

This returns a 'FARS' object: a list with six tibbles: `flat`, `multi_acc`, 
`multi_veh`, `multi_per`, `events`, and `codebook`.

The `flat` tibble contains all variables for which there is just one value per
crash ("accident"), vehicle, or person (e.g., weather conditions, travel speed, 
age). Each row corresponds to a person involved in a crash. As there may be 
multiple people and/or vehicles involved in one crash, some variable-values are 
repeated within a crash or vehicle. Each crash is uniquely identified with `id`,
which is a combination of `year` and `st_case`. Note that `st_case` is not 
unique across years, for example, `st_case` 510001 will appear in each year. 
The `id` variable attempts to avoid this issue.

Please review the [FARS Analytical User's Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813417) for more information.

```{r}
glimpse(myFARS$flat)
```

The `multi_` tibbles contain those variables for which there may be a varying
number of values for any entity (e.g., driver impairments, vehicle events,
weather conditions at time of crash). Each tibble has the requisite data
elements corresponding to the entity: `multi_acc` includes `st_case` and `year`,
`multi_veh` adds `veh_no` (vehicle number), and `multi_per` adds `per_no`
(person number). 

The top name-value pairs of each tibble are shown below.

```{r}
bind_rows(
  
  myFARS$multi_acc %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% mutate(table = "multi_acc"),
  myFARS$multi_veh %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% mutate(table = "multi_veh"),
  myFARS$multi_per %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% mutate(table = "multi_per"),
  
) %>%
  
  select(table, name, value, n)
```


The `events` tibble provides a sequence of numbered events for each vehicle in
each crash.

```{r}
head(myFARS$events, 10)
```


The `codebook` tibble provides a searchable codebook for the data, useful if you
know what concept you're looking for but not the variable that describes it.

```{r}
# View(myFARS$codebook)

myFARS$codebook %>%
  filter(name_rfars == "lgt_cond") %>% distinct() %>% head(10)
```


### Counts

A first step in many transportation safety analyses involves counting the number
of relevant crashes, fatalities, or people involved. `counts()` lets users
specify *what* to count, *where* to count them (rural/urban and/or in specified
states or regions), *who* to include, the *interval* over which to count 
(annually or monthly), and factors *involved* in the crashes. It
returns a simple tibble that can be easily piped into `ggplot()` to quickly
visualize counts.

```{r, fig.height=2.2}
my_counts <- counts(
  myFARS,
  what = "crashes",
  interval = c("year")
  ) 

head(my_counts)

my_counts %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatal Crashes in Virginia")
```

```{r, fig.height=2.2}
counts(
  myFARS,
  what = "fatalities",
  interval = c("year")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       where = list(urb="rural"),
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Rural Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       where = list(urb="rural"),
       interval = c("year"),
       involved = "speeding"
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Rural Virginia")
```

We can combine two `counts()` results to make a comparison. Here we compare the
number of speeding-related fatalities in rural and urban Virginia:

```{r, fig.height=2.4}
compare_counts(
  df = myFARS,
  involved = "speeding",
  what = "fatalities",
  where = list(urb="rural"),
  where2 = list(urb="urban")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~urb) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

### Mapping

We can take advantage of having access to the full data with maps. Here we map
pedestrian and bicyclist fatalities in Virginia:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "pedbike", 
  filterOnly = TRUE
  ) %>% 
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

Drug-related crashes:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "drugs", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  filter(lat < 90, lon < 900) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

Young drivers:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "young driver", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  filter(lat < 90, lon < 900) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

### Modeling

Having access to the full dataset also allows us to develop statistical models.
Here we fit a simple model of injury severity as a function of age and restraint
use. The results indicate that injury severity increases with age, and when 
seat belts are not used properly. 

```{r, results='asis'}
# table(myFARS$flat$inj_sev)
# table(myFARS$flat$rest_use, useNA = "ifany")
# table(myFARS$flat$per_typ, useNA = "ifany")

model_data <-
  myFARS$flat %>%
  filter(rest_use %in% c("Lap Belt Only Used",
                         "Shoulder Belt Only Used",
                         "None Used / Not Applicable",
                         "None Used/Not Applicable",
                         "Shoulder and Lap Belt Used"),
         per_typ %in% c("Driver of a Motor Vehicle In-Transport",
                        "Passenger of a Motor Vehicle In-Transport")
         ) %>%
  mutate(
    rest_use = case_when(
      rest_use %in% c("Lap Belt Only Used", "Shoulder Belt Only Used") ~ "Partial",
      rest_use %in% c("None Used / Not Applicable", "None Used/Not Applicable") ~ "None",
      rest_use %in% c("Shoulder and Lap Belt Used") ~ "Full",
      TRUE ~ "Unknown"
      ) %>%
      as.factor() %>%
      relevel(ref = "Full"),
    kabco = case_when(
      inj_sev == "Fatal Injury (K)" ~ 4,
      inj_sev %in% c("Suspected Serious Injury (A)", 
                     "Suspected Serious Injury(A)") ~ 3,
      inj_sev %in% c("Suspected Minor Injury (B)", 
                     "Suspected Minor Injury(B)") ~ 2,
      inj_sev == "Possible Injury (C)" ~ 1,
      inj_sev == "No Apparent Injury (O)" ~ 0,
      TRUE ~ as.numeric(NA)
      ),
    age_n = gsub("\\D+","", age) %>% as.numeric()) 

my_model <- lm(kabco ~ age_n + rest_use, data = model_data) 

stargazer::stargazer(my_model, type = "html")
```


```{r}
new_data <- expand.grid(
  age_n = c(20, 60),
  rest_use = factor(c("Full", "Partial", "None"), levels = c("Full", "Partial", "None"), ordered = TRUE) )

new_data %>%
  mutate(pred = predict(my_model, newdata = new_data),
         age = paste0(age_n, " yrs")) %>%
  ggplot(aes(x=rest_use, y=pred)) +
    geom_col() +
    facet_wrap(.~age) +
    scale_y_continuous(
      limits = c(0,4), breaks = 0:4, labels = c("O", "C", "B", "A", "K"), expand = expansion()) +
    theme(panel.grid = element_blank(),
          panel.grid.major.y = element_line(c("black")),
          axis.ticks = element_blank()) +
    labs(x="", y="", title = "Predicted Injury Severity by Age and Restraint Use",
         caption = "Full = correctly used seatbelt, partial = partially correctly used, none = no seatbelt.")
```

```{r, include=FALSE}
#unlink(paste0(getwd(),"/FARS data"), recursive = TRUE)
```


## Getting and Using GES/CRSS Data

Use the `get_grscrss()` function to bring GES/CRSS data into the current environment and optionally
save an RDS file for future use (faster than downloading anew). 
This is done by (a) downloading the data to a temporary directory, (b) downloading
to a permanent directory, or (c) importing from a permanent directory. After 
data is downloaded to a permanent directory, the function will look there rather
than downloading the data again. If a year of data is requested but not found,
R will ask your permission to download the missing data. 

Here we get three years of data. We do not specify a region so that we can 
confirm the results with official data.

```{r}
myGESCRSS <- get_gescrss(years = 2019:2021)
```

This returns a 'GESCRSS' object: a list with six tibbles: `flat`, `multi_acc`, 
`multi_veh`, `multi_per`, `events`, and `codebook`.

The `flat` tibble contains all variables for which there is just one value per
crash ("accident"), vehicle, or person (e.g., weather conditions, travel speed, 
age). Each row corresponds to a person involved in a crash. As there may be 
multiple people and/or vehicles involved in one crash, some variable-values are 
repeated within a crash or vehicle. Each crash is uniquely identified with `id`,
which is a combination of `year` and `casenum`. 

Please review the [CRSS Analytical User's Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813436) for more information.

```{r}
glimpse(myGESCRSS$flat)
```

The `multi_` tibbles contain those variables for which there may be a varying
number of values for any entity (e.g., driver impairments, vehicle events,
weather conditions at time of crash). Each tibble has the requisite data
elements corresponding to the entity: `multi_acc` includes `st_case` and `year`,
`multi_veh` adds `veh_no` (vehicle number), and `multi_per` adds `per_no`
(person number). 

The top three name-value pairs of each tibble are shown below.

```{r}
myGESCRSS$multi_acc %>% group_by(name, value) %>% summarize(n=n()) %>% 
  arrange(name, desc(n)) %>% mutate(rank=row_number()) %>%
  filter(rank<=3) %>%
  select(-rank)

myGESCRSS$multi_veh %>% group_by(name, value) %>% summarize(n=n()) %>% 
  arrange(name, desc(n)) %>% mutate(rank=row_number()) %>%
  filter(rank<=3) %>%
  select(-rank)

myGESCRSS$multi_per %>% group_by(name, value) %>% summarize(n=n()) %>% 
  arrange(name, desc(n)) %>% mutate(rank=row_number()) %>%
  filter(rank<=3) %>%
  select(-rank)
```


The `events` tibble provides a sequence of numbered events for each vehicle in
each crash.

```{r}
head(myGESCRSS$events, 10)
```


The `codebook` tibble provides a searchable codebook for the data, useful if you
know what concept you're looking for but not the variable that describes it.

```{r}
# View(myGESCRSS$codebook)

myGESCRSS$codebook %>%
  filter(name_rfars == "lgt_cond") %>% distinct() %>% head(10)
```


### Counts

A first step in many transportation safety analyses involves counting the number
of relevant crashes, fatalities, or people involved. `counts()` lets users
specify *what* to count, *where* to count them (rural/urban and/or in specified
states or regions), *who* to include, which *years* and an aggregation
*interval* (annually or monthly), and factors *involved* in the crash. It
returns a simple tibble that can be easily piped into `ggplot()` to quickly
visualize counts.

```{r, fig.height=2.2}
my_counts <- counts(
  myGESCRSS,
  what = "crashes",
  interval = c("year")
  ) 

head(my_counts)

my_counts %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Total Estimated Crashes")
```


```{r, fig.height=2.2}
counts(
  myGESCRSS,
  where = list(region="s"),
  what = "injuries",
  interval = c("year")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Injuries in the South")
```



```{r, fig.height=2.4}
compare_counts(
  df = myGESCRSS,
  what = "injuries",
  where = list(region="s"),
  where2 = list(region="ne")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~region) +
    labs(x=NULL, y=NULL, title = "Injuries in the South and Northeast", fill=NULL)
```

### Mapping

GES/CRSS is not very friendly to mapping. It does not provide lat/lon 
coordinates or States. Regions are composed of several States and require extra
steps to generate mappable data.


### Modeling

Having access to the full dataset also allows us to develop statistical models.
Here we fit a simple model of injury severity as a function of age and restraint
use. The results indicate that injury severity increases with age, and when 
seat belts are not used properly. 

```{r, results='asis'}
# table(myFARS$flat$inj_sev)
# table(myFARS$flat$rest_use, useNA = "ifany")
# table(myFARS$flat$per_typ, useNA = "ifany")

model_data <-
  myGESCRSS$flat %>%
  filter(rest_use %in% c("Lap Belt Only Used",
                         "Shoulder Belt Only Used",
                         "None Used / Not Applicable",
                         "None Used/Not Applicable",
                         "Shoulder and Lap Belt Used"),
         per_typ %in% c("Driver of a Motor Vehicle In-Transport",
                        "Passenger of a Motor Vehicle In-Transport")
         ) %>%
  mutate(
    rest_use = case_when(
      rest_use %in% c("Lap Belt Only Used", "Shoulder Belt Only Used") ~ "Partial",
      rest_use %in% c("None Used / Not Applicable", "None Used/Not Applicable") ~ "None",
      rest_use %in% c("Shoulder and Lap Belt Used") ~ "Full",
      TRUE ~ "Unknown"
      ) %>%
      as.factor() %>%
      relevel(ref = "Full"),
    kabco = case_when(
      inj_sev == "Fatal Injury (K)" ~ 4,
      inj_sev %in% c("Suspected Serious Injury (A)", 
                     "Suspected Serious Injury(A)") ~ 3,
      inj_sev %in% c("Suspected Minor Injury (B)", 
                     "Suspected Minor Injury(B)") ~ 2,
      inj_sev == "Possible Injury (C)" ~ 1,
      inj_sev == "No Apparent Injury (O)" ~ 0,
      TRUE ~ as.numeric(NA)
      ),
    age_n = gsub("\\D+","", age) %>% as.numeric(),
    speed_n = gsub("\\D+","", trav_sp) %>% as.numeric()
    ) 

my_model <- lm(kabco ~ age_n + rest_use + speed_n, data = model_data) 

stargazer::stargazer(my_model, type = "html")
```


```{r}
new_data <- expand.grid(
  age_n = c(20, 60),
  speed_n = seq(10, 60, 10),
  rest_use = factor(c("Full", "Partial", "None"), levels = c("Full", "Partial", "None"), ordered = TRUE) )

new_data %>%
  mutate(pred = predict(my_model, newdata = new_data),
         age = paste0(age_n, " yrs")) %>%
  ggplot(aes(x=speed_n, y=pred, fill=age)) +
    geom_col(position = position_dodge()) +
    facet_wrap(.~rest_use) +
    scale_y_continuous(
      limits = c(0,4), breaks = 0:4, labels = c("O", "C", "B", "A", "K"), expand = expansion()) +
    theme(panel.grid = element_blank(),
          panel.grid.major.y = element_line(c("black")),
          axis.ticks = element_blank()) +
    labs(x="Travel Speed (mph)", y="", title = "Predicted Injury Severity by Age, Restraint Use, and Travel Speed",
         fill="Age",
         caption = "Full = correctly used seatbelt, partial = partially correctly used, none = no seatbelt.")
```

```{r, include=FALSE}
unlink(paste0(getwd(),"/FARS data"), recursive = TRUE)
```
