---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi = 500
)
```

# rfars <img src="man/figures/logo.svg" align="right" alt="" width="120" />

<!-- badges: start -->

[![Travis build
status](https://travis-ci.com/s87jackson/rfars.svg?branch=master)](https://app.travis-ci.com/github/s87jackson/rfars)

<!-- badges: end -->

The goal of `rfars` is to simplify the process of analyzing FARS data. FARS
stands for [Fatality Analysis Reporting
System](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars),
the census of fatal crashes in the United States maintained by the [National
Highway Traffic Safety Administration](https://www.nhtsa.gov/). The [Fatality
and Injury Reporting System Tool](https://cdan.dot.gov/query) allows users to
generate queries, and can produce simple tables and graphs. This suffices for
simple analysis, but often leaves researchers wanting more. Digging any deeper,
however, involves a time-consuming process of downloading annual ZIP files and
attempting to stitch them together - after first combing through the immense
[data
dictionary](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813254)
to determine the required variables and table names. `rfars` allows users to
download FARS data back to 2015 with just one line of code. The result is a
full, rich dataset ready for mapping, modeling, and other downstream analysis.
Helper functions are also provided to produce common counts and comparisons.

A companion package `rfarsplus` (currently in development) will provide exposure
data and facilitate the calculation of various rates.

## Installation

You can install the latest version of `rfars` from [GitHub](https://github.com/)
with:

``` r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

Then load the required packages:

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(rfars)
library(dplyr)
library(ggplot2)
library(leaflet)
library(leaflet.extras)
```

## Getting and Using FARS Data

Use the `get_fars()` function to bring FARS data into the current environment. 
This is done by (a) downloading the data to a temporary directory, (b) downloading
to a permanent directory, or (c) importing from a permanent directory. After 
data is downloaded to a permanent directory, the function will look there rather
than downloading the data again. If a year of data is requested but not found,
R will ask your permission to download the missing data. 

Here we get three years of data for Virginia:

```{r, warning=FALSE}
myFARS <- get_fars(years = 2018:2020, states = "VA")
```

This returns a 'FARS' object: a list with five tibbles: `flat`, `multi_acc`, 
`multi_veh`, `multi_per`, and `events`.

The `flat` tibble contains all variables for which there is just one value per
crash ("accident"), vehicle, or person (e.g., weather conditions, travel speed, 
age). Each row corresponds to a person involved in a crash. As there may be 
multiple people and/or vehicles involved in one crash, some variable-values are 
repeated within a crash or vehicle. Each crash is uniquely identified with `id`,
which is a combination of `year` and `st_case`. Note that `st_case` is not 
unique across years, for example, `st_case` 510001 will appear in each year. 
The `id` variable attempts to avoid this issue.

The `multi_` tibbles contain those variables for which there may be a varying
number of values for any entity (e.g., driver impairments, vehicle events,
weather conditions at time of crash). Each tibble has the requisite data
elements corresponding to the entity: `multi_acc` includes `st_case` and `year`,
`multi_veh` adds `veh_no` (vehicle number), and `multi_per` adds `per_no`
(person number).

The `events` tibble provides a sequence of numbered events for each vehicle in
each crash.

```{r}
glimpse(myFARS)
```

You can review the list of variables to help guide your analysis with:

``` r
View(fars_varnames)
```

## Counts

A first step in many transportation safety analyses involves counting the number
of relevant crashes, fatalities, or people involved. `counts()` lets users
specify *what* to count, *where* to count them (rural/urban and/or in specified
states), *who* to include, which *years* to include and an aggregation
*interval* (annually or monthly), and factors *involved* in the crash. It
returns a simple tibble that can be easily piped into `ggplot()` to quickly
visualize counts.

```{r, fig.height=2.2}
counts(
  myFARS,
  what = "crashes",
  interval = c("year")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatal Crashes in Virginia")
```

```{r, fig.height=2.2}
counts(
  myFARS,
  what = "fatalities",
  interval = c("year")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       where = "rural",
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Rural Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Rural Virginia")
```

We can combine two `counts()` results to make a comparison. Here we compare the
number of speeding-related fatalities in rural and urban Virginia:

```{r, fig.height=2.4}
bind_rows(
  counts(myFARS,
       what = "fatalities",
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Rural"),
  counts(myFARS,
       what = "fatalities",
       where = "urban",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Urban")
    ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~where) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

## Mapping

We can take advantage of having access to the full data with maps. Here we map
pedestrian and bicyclist fatalities in Virginia:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "pedbike", 
  filterOnly = TRUE
  ) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

Drug-related crashes:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "drugs", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

Young drivers:

```{r}
counts(
  myFARS, 
  what = "crashes", 
  involved = "young driver", 
  filterOnly = TRUE
  ) %>%
  filter(!is.na(lat), !is.na(lon)) %>%
leaflet() %>%
  addTiles() %>%
  addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 1) %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7, group = "Crash Locations") 
```

## Modeling

Having access to the full dataset also allows us to develop statistical models.
Here we fit a simple model of injury severity as a function of age and restraint
use. The results indicate that injury severity increases with age, and when 
seat belts are not used properly. 

```{r, results='asis'}
# table(myFARS$flat$inj_sev)
# table(myFARS$flat$rest_use, useNA = "ifany")
# table(myFARS$flat$per_typ, useNA = "ifany")

model_data <-
  myFARS$flat %>%
  filter(rest_use %in% c("Lap Belt Only Used",
                         "Shoulder Belt Only Used",
                         "None Used / Not Applicable",
                         "None Used/Not Applicable",
                         "Shoulder and Lap Belt Used"),
         per_typ %in% c("Driver of a Motor Vehicle In-Transport",
                        "Passenger of a Motor Vehicle In-Transport")
         ) %>%
  mutate(
    rest_use = case_when(
      rest_use %in% c("Lap Belt Only Used", "Shoulder Belt Only Used") ~ "Partial",
      rest_use %in% c("None Used / Not Applicable", "None Used/Not Applicable") ~ "None",
      rest_use %in% c("Shoulder and Lap Belt Used") ~ "Full",
      TRUE ~ "Unknown"
      ) %>%
      as.factor() %>%
      relevel(ref = "Full"),
    kabco = case_when(
      inj_sev == "Fatal Injury (K)" ~ 4,
      inj_sev %in% c("Suspected Serious Injury (A)", 
                     "Suspected Serious Injury(A)") ~ 3,
      inj_sev %in% c("Suspected Minor Injury (B)", 
                     "Suspected Minor Injury(B)") ~ 2,
      inj_sev == "Possible Injury (C)" ~ 1,
      inj_sev == "No Apparent Injury (O)" ~ 0,
      TRUE ~ as.numeric(NA)
      ),
    age_n = gsub("\\D+","", age) %>% as.numeric()) 

my_model <- lm(kabco ~ age_n + rest_use, data = model_data) 

stargazer::stargazer(my_model, type = "html")
```


```{r}
new_data <- expand.grid(
  age_n = c(20, 60),
  rest_use = factor(c("Full", "Partial", "None"), levels = c("Full", "Partial", "None"), ordered = TRUE) )

new_data %>%
  mutate(pred = predict(my_model, newdata = new_data),
         age = paste0(age_n, " yrs")) %>%
  ggplot(aes(x=rest_use, y=pred)) +
    geom_col() +
    facet_wrap(.~age) +
    scale_y_continuous(
      limits = c(0,4), breaks = 0:4, labels = c("O", "C", "B", "A", "K"), expand = expansion()) +
    theme(panel.grid = element_blank(),
          panel.grid.major.y = element_line(c("black")),
          axis.ticks = element_blank()) +
    labs(x="", y="", title = "Predicted Injury Severity by Age and Restraint Use",
         caption = "Full = correctly used seatbelt, partial = partially correctly used, none = no seatbelt.")
```

```{r, include=FALSE}
unlink(paste0(getwd(),"/FARS data"), recursive = TRUE)
```
