---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rfars

<!-- badges: start -->

<!-- badges: end -->

The goal of rfars is to simplify the process of analyzing FARS data. The
[Fatality and Injury Reporting System Tool](https://cdan.dot.gov/query) allows
users to generate queries, and can produce simple tables and graphs. This
suffices for simple analysis, but often leaves researchers wanting more. Digging
any deeper, however, involves a time-consuming process of downloading annual ZIP
files and attempting to stitch them together - after first combing through the
immense [data
dictionary](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813254)
to determine the required variables and table names. `rfars`allows users to
download five years of FARS data with just two lines of code. The result is a
full, rich dataset ready for mapping, modeling, and other downstream analysis.
Helper functions are also provided to produce common counts and comparisons. A
companion package `rfarsplus`provides exposure data and facilitates the
calculation of various rates.

## Installation

You can install the latest version of `rfars`from [GitHub](https://github.com/)
with:

``` r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

## Getting Data

Here we import 5 years of data for Virginia. Note that `get_fars`requires your
permission to download the ZIP files from NHTSA and save the prepared files on
your hard drive. This only has to be run once and defaults to saving everything
in the current working directory. The `use_fars`function looks in that directory
for certain files and compiles them into a list of data frames: `flat`,
`multi_acc`, `multi_veh`, and `multi_per`.

```{r example}
library(rfars)
library(dplyr)

# get_fars(years = 2016:2020, states="Virginia")

myFARS <- use_fars() 

#head(myFARS)
```

You can review the list of variables to help guide your analysis with:

``` r
View(fars_varnames)
```

## Counts

A first step in many transportation analyses involves counting the number of
relevant crashes, fatalities, or people involved. `counts`lets users specify a
time period and aggregation interval, and focus in on specific road users and
factors. It can be combined with `ggplot`to quickly visualize counts.

```{r, fig.height=2.2}
library(ggplot2)

counts(myFARS,
       what = "crashes",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Crashes in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Fatalities in Virginia")
```

```{r, fig.height=2.2}
counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year")
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Annual Rural Fatalities in Virginia")
```

We can combine two `counts`to make a comparison. Here we compare the number of
speeding-related fatalities in rural and urban places:

```{r, fig.height=2.4}
bind_rows(
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "rural",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Rural"),
  counts(myFARS,
       what = "fatalities",
       when = 2016:2020,
       where = "urban",
       interval = c("year"),
       involved = "speeding"
       ) %>%
    mutate(where = "Urban")
    ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~where) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

## Mapping

We can take advantage of having access to the full data with maps. Here we map
the locations of pedbikes involved in fatal crashes in Virginia:

```{r}
library(leaflet)
library(leaflet.extras)

myFARS$flat %>% 
  filter(per_typ %in% c("Bicyclist", "Pedestrian")) %>%
  #filter(per_typ %in% c("Bicyclist")) %>%
  leaflet() %>%
    addTiles() %>%
    addHeatmap(group = "Heatmap", radius=10, blur=20, minOpacity = .01, max = .2, cellSize = 10) %>%
    addCircleMarkers(
      radius = 1,
      color = "red",
      stroke = FALSE,
      fillOpacity = 0.7, group = "Crash Locations") %>%
    addLayersControl(
      overlayGroups = c("Crash Locations", "Heatmap"),
      options = layersControlOptions(collapsed = FALSE))
```

We can also use the `counts(..., filterOnly=TRUE)` to access the filtered data
prior to aggregating:

```{r}
counts(myFARS, what = "fatalities", when = 2016:2020, involved = "alcohol", 
       filterOnly = TRUE) %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 1,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.7)
```

## Modeling

Having access to the full dataset also allows us to develop statistical models.
Here we fit a simple model of injury severity:

```{r}
table(myFARS$flat$inj_sev)
table(myFARS$flat$rest_mis)

myFARS$flat %>%
  mutate(kabco = case_when(inj_sev == "Fatal Injury (K)" ~ 4,
                           inj_sev %in% c("Suspected Serious Injury (A)", 
                                          "Suspected Serious Injury(A)") ~ 3,
                           inj_sev %in% c("Suspected Minor Injury (B)", 
                                          "Suspected Minor Injury(B)") ~ 2,
                           inj_sev == "Possible Injury (C)" ~ 1,
                           inj_sev == "No Apparent Injury (O)" ~ 0,
                           TRUE ~ as.numeric(NA)
                           ),
         age_n = gsub("\\D+","", age) %>% as.numeric()) %>%
  lm(kabco ~ age_n + rest_mis, data = .) %>%
  summary()
```
