---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%", fig.align = "center",
  dpi = 500
)
```

# rfars <img src="man/figures/logo.svg" align="right" width="120"/>

<!-- badges: start -->

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/rfars)](https://cran.r-project.org/package=rfars)
[![R CMD
Check](https://github.com/s87jackson/rfars/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/s87jackson/rfars/actions/workflows/R-CMD-check.yaml)
[![](https://cranlogs.r-pkg.org/badges/grand-total/rfars)](https://CRAN.R-project.org/package=rfars)

<!-- badges: end -->

The goal of `rfars` is to facilitate transportation safety analysis by
simplifying the process of extracting data from official crash databases. The
[National Highway Traffic Safety Administration](https://www.nhtsa.gov/)
collects and publishes a census of fatal crashes in the [Fatality Analysis
Reporting
System](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars)
and a sample of fatal and non-fatal crashes in the [Crash Report Sampling
System](https://www.nhtsa.gov/crash-data-systems/crash-report-sampling-system)
(an evolution of the [General Estimates
System](https://www.nhtsa.gov/national-automotive-sampling-system/nass-general-estimates-system)).
The [Fatality and Injury Reporting System Tool](https://cdan.dot.gov/query)
allows users to query these databases, and can produce simple tables and graphs.
This suffices for simple analysis, but often leaves researchers wanting more.
Digging any deeper, however, involves a time-consuming process of downloading
annual ZIP files and attempting to stitch them together - after first combing
through immense data dictionaries to determine the required variables and table
names.

`rfars` allows users to download FARS and GES/CRSS data back to 2011 with just
one line of code. The result is a full, rich dataset ready for mapping,
modeling, and other downstream analysis. Helper functions are also provided to
produce common counts and comparisons.

## Installation

You can install the latest version of `rfars` from [GitHub](https://github.com/)
with:

``` r
# install.packages("devtools")
devtools::install_github("s87jackson/rfars")
```

or the CRAN stable release with:

``` r
install.packages("rfars")
```

Then load rfars and some helpful packages:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(rfars)
library(dplyr)
```

## Getting and Using Data

The `get_fars()` and `get_gescrss()` are the primary functions of the `rfars`
package. These functions either download and process data files directly from
[NHTSA's FTP Site](https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/), or
pull the prepared file stored on your local machine. They take the functions
`years` and `states` (FARS) or `regions` (GES/CRSS). As the source data files
follow an annual structure, `years` determines how many file sets are
downloaded, and `states`/`regions` filters the resulting dataset. Downloading
and processing these files can take several minutes. Before downloading `rfars`
will inform you that it's about to download files and asks your permission to do
so. To skip this dialog, set `proceed = TRUE`. You can use the `dir` and `cache`
parameters to save an RDS file to your local machine. The `dir` parameter
specifices the directory, and `cache` names the file (be sure to include the
.rds file extension).

Here we get one year of FARS data for Virginia:

```{r}
myFARS <- get_fars(years = 2022, states = "VA", proceed = TRUE)
```

We could have saved that file locally with:

```{r, eval=F}
myFARS <- get_fars(years=2022, states = "VA", proceed = TRUE, dir = getwd(), cache = "myFARS.rds")
```

Note that you can assign and save this data with one function call.

We could similarly get one year of CRSS data for the south (MD, DE, DC, WV, VA,
KY, TN, NC, SC, GA, FL, AL, MS, LA, AR, OK, TX):

```{r, eval=F}
myCRSS <- get_gescrss(years = 2022, regions = "s", proceed = TRUE)
myCRSS <- get_gescrss(years = 2022, regions = "s", proceed = TRUE, dir = getwd(), cache = "myCRSS.rds")
```

The data returned by `get_fars()` and `get_gescrss()` adhere to the same
structure: a list with six tibbles: `flat`, `multi_acc`, `multi_veh`,
`multi_per`, `events`, and `codebook`. FARS and GES/CRSS share many but not all
data elements. See the [FARS Analytical User's
Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813417) and
[CRSS Analytical User's
Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813436) for
more information.

The following section decribes the results from `get_fars()` but largely applies
to `get_gescrss()` as well.

The `flat` tibble contains all variables for which there is just one value per
crash ("accident"), vehicle, or person (e.g., intersection type, travel speed,
age). Each row corresponds to a person involved in a crash. As there may be
multiple people and/or vehicles involved in one crash, some variable-values are
repeated within a crash or vehicle. Each crash is uniquely identified with `id`,
which is a combination of `year` and `st_case`. Note that `st_case` is not
unique across years, for example, `st_case` 510001 will appear in each year. The
`id` variable attempts to avoid this issue.

```{r}
glimpse(myFARS$flat, width = 100)
```

The `multi_` tibbles contain those variables for which there may be a varying
number of values for any entity (e.g., driver impairments, vehicle events,
weather conditions at time of crash). Each tibble has the requisite data
elements corresponding to the entity: `multi_acc` includes `st_case` and `year`,
`multi_veh` adds `veh_no` (vehicle number), and `multi_per` adds `per_no`
(person number).

The top name-value pairs of each tibble are shown below.

```{r, results='asis'}
myFARS$multi_acc %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% select(name, value, n) %>% knitr::kable(format = "html")
```

```{r, results='asis'}
myFARS$multi_veh %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% select(name, value, n) %>% knitr::kable(format = "html")
```

```{r, results='asis'}
myFARS$multi_per %>% filter(!is.na(value)) %>% group_by(name, value) %>% summarize(n=n(), .groups = "drop") %>% arrange(desc(n)) %>% slice(1:10) %>% select(name, value, n) %>% knitr::kable(format = "html")
```

The `events` tibble provides a sequence of numbered events for each vehicle in
each crash. See the vignette for more information.

```{r, results='asis'}
head(myFARS$events, 10) %>% knitr::kable(format="html")
```

The `codebook` tibble provides a searchable codebook for the data, useful if you
know what concept you're looking for but not the variable that describes it. The
`rfars` package includes a codebook for FARS and GESCRSS (`rfars::fars_codebook`
and `rfars::gescrss_codebook`). These tables span 2011-2022 whereas the
`codebook` object returned from `get_fars()` and `get_gescrss()` only include
the specified `years`. See the vignette for more information.

## Helpful Links

-   [National Highway Traffic Safety Administration
    (NHTSA)](https://www.nhtsa.gov/)
-   [Fatality Analysis Reporting System
    (FARS)](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars)
-   [Fatality and Injury Reporting System Tool
    (FIRST)](https://cdan.dot.gov/query)
-   [FARS Analytical User's
    Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813417)
-   [General Estimates System
    (GES)](https://www.nhtsa.gov/national-automotive-sampling-system/nass-general-estimates-system)
-   [Crash Report Sampling System
    (CRSS)](https://www.nhtsa.gov/crash-data-systems/crash-report-sampling-system)
-   [CRSS Analytical User's
    Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813436)
-   [NCSA and Other Data
    Sources](https://cdan.dot.gov/Homepage/MotorVehicleCrashDataOverview.htm)
-   [NHTSA FTP Site](https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/)

```{r, include=FALSE}
unlink(paste0(getwd(),"/FARS data"), recursive = TRUE)
```
