---
title: "Counts"
output: rmarkdown::html_vignette
fig.width: 8
fig.height: 5
vignette: >
  %\VignetteIndexEntry{Counts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

## Counts

A first step in many transportation safety analyses involves counting the number
of relevant crashes, fatalities, or people involved. `counts()` lets users
specify *what* to count, *where* to count them (rural/urban and/or in specified
states or regions), *who* to include, the *interval* over which to count 
(annually or monthly), and factors *involved* in the crashes. It
returns a simple tibble that can be easily piped into `ggplot()` to quickly
visualize counts.

### Generating Counts

First we load the required libraries:

```{r, message=F}
library(rfars)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Then pull a year of FARS data for Virginia:

```{r}
myFARS <- get_fars(years = 2021, states = "VA", proceed = T)
```

Then we can use `counts()` to reduce the data to desired counts.

Here we count crashes:

```{r, results='asis'}
my_counts <- counts(
  myFARS,
  what = "crashes",
  interval = c("month")
  )
```

This returns the following dataframe:

```{r, results='asis'}
knitr::kable(my_counts, format = "html")
```

Which we can graph:

```{r}
my_counts %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatal Crashes in Virginia")
```

We could alternatively count fatalities:

```{r}
counts(
  myFARS,
  what = "fatalities",
  interval = c("month")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Fatalities in Virginia")
```

Or fatalities involving speeding:

```{r}
counts(myFARS,
       what = "fatalities",
       interval = c("month"),
       involved = "speeding"
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia")
```

Or fatalities involving speeding in rural areas:

```{r}
counts(myFARS,
       what = "fatalities",
       where = list(urb="rural"),
       interval = c("month"),
       involved = "speeding"
       ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Rural Virginia")
```


### Comparing Counts

We can use `compare_counts()` to quickly produce comparison graphs.

Here we compare speeding-related fatalities in rural and urban areas:

```{r}
compare_counts(
  df = myFARS,
  interval = "month",
  involved = "speeding",
  what = "fatalities",
  where = list(urb="rural"),
  where2 = list(urb="urban")
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~urb) +
    labs(x=NULL, y=NULL, title = "Speeding-Related Fatalities in Virginia", fill=NULL)
```

And here we compare speeding-related crashes to those related to distraction:

```{r}
compare_counts(
  df = myFARS,
  interval = "month",
  involved = "speeding",
  involved2 = "distracted driver",
  what = "crashes",
  ) %>%
  ggplot(aes(x=date, y=n, label=scales::comma(n))) + 
    geom_col() + 
    geom_label(vjust=1.2) +
    facet_wrap(.~involved) +
    labs(x=NULL, y=NULL, title = "Speeding- and Distraction-Related Crashes in Virginia", fill=NULL)
```

### annual_counts data table

*rfars* includes **annual_counts**, a table of annual crash counts and 
related metrics:

```{r, fig.height=5, fig.width=12.5, dpi=800}
rfars::annual_counts %>%
  filter(involved != "general") %>%
  replace_na(list(type="")) %>%
  unite(type, what, col="what2", sep = " ", remove = T) %>%
  mutate_at(c("involved", "what2"), stringr::str_to_sentence) %>%

  ggplot(aes(x=year, y=n)) +
  geom_line() +
  facet_grid(what2~involved, scales="free_y", labeller = label_wrap_gen(width = 10)) +
  scale_y_continuous(label=scales::comma, expand = expansion()) +
  scale_x_continuous(labels = seq(2014, 2023, 2), breaks = seq(2014, 2023, 2), expand = expansion()) +
  labs(x=NULL, y=NULL, title = "Crash Types Over Time") +
  theme(
    strip.text.y.right = element_text(angle=0),
    axis.ticks = element_blank(),
    axis.text.x.bottom = element_text(angle=45, hjust=1, size=8)
    )
```


```{r}
rfars::annual_counts %>%
  filter(involved == "general") %>%
  replace_na(list(type="")) %>%
  unite(type, what, col="what2", sep = " ", remove = T) %>%
  mutate_at(c("involved", "what2"), stringr::str_to_sentence) %>%

  ggplot(aes(x=year, y=n)) +
  geom_line() +
  facet_grid(what2~., scales="free_y", labeller = label_wrap_gen(width = 10)) +
  scale_y_continuous(label=scales::comma, expand = expansion()) +
  scale_x_continuous(labels = seq(2014, 2023, 2), breaks = seq(2014, 2023, 2), expand = expansion()) +
  labs(x=NULL, y=NULL, title = "Crash Types Over Time") +
  theme(
    strip.text.y.right = element_text(angle=0),
    axis.ticks = element_blank()
  )


rfars::annual_counts %>%
  filter(involved %in% c("pedalcyclist", "bicyclist", "pedestrian", "pedbike")) %>%
  replace_na(list(type="")) %>%
  unite(type, what, col="what2", sep = " ", remove = T) %>%
  mutate_at(c("involved", "what2"), stringr::str_to_sentence) %>%

  ggplot(aes(x=year, y=n)) +
  geom_line() +
  facet_grid(what2~involved, scales="free_y", labeller = label_wrap_gen(width = 10)) +
  scale_y_continuous(label=scales::comma, expand = expansion()) +
  scale_x_continuous(labels = seq(2014, 2023, 2), breaks = seq(2014, 2023, 2), expand = expansion()) +
  labs(x=NULL, y=NULL, title = "Crash Types Over Time") +
  theme(
    strip.text.y.right = element_text(angle=0),
    axis.ticks = element_blank(),
    axis.text.x.bottom = element_text(angle=45, hjust=1, size=8)
  )
```
