---
title: "Crash Sequence of Events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crash Sequence of Events}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

This vignette explores the **vsoe** (Vehicle Sequence of Events) data to
visualize crash sequence patterns.

**vsoe** is one of three event-based data files, the others being **cevent** and
**vevent**. According to the [CRSS Analytical User's Manual](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813707),
**vevent** "has the same data elements as the **cevent** data file" plus "a data
element that records the sequential event number for each vehicle," and the
**vsoe** file "has a subset of the data elements contained in the **Vevent**
data file (it is a simplified **vevent** data file)" (p. 16). `rfars` therefore
omits **cevent** and **vevent**.

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(rfars)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyr)
```


First we get one year of data, and filter to the southern region for simplicity. 

```{r, warning=FALSE}
mydata <- rfars::get_gescrss(years=2023, regions = "s", proceed = TRUE)

my_events <- mydata$events
```

The **Vsoe** data is stored in the `events` tibble of the object returned by
`get_gescrss()`. Here we see the top 10 individual events:

```{r, results='asis'}
my_events %>%
  group_by(soe) %>% summarize(n=n()) %>%
  arrange(desc(n)) %>%
  slice(1:10) %>%
  knitr::kable(format = "html")
```

We can also see the top 10 most common sequences:

```{r, results='asis'}
my_events %>%
  select(-aoi) %>%
  mutate_at("soe", as.character) %>%
  pivot_wider(
    names_from = "veventnum", 
    values_from = "soe", 
    values_fill = "x",
    names_prefix = "event") %>%
  select(starts_with("event")) %>%
  group_by_all() %>%
  summarize(n=n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  slice(1:10) %>%
  select(event1, event2, n) %>%
  knitr::kable(format = "html")
```

Below we consider all state transitions - the transition from one event to the
next in the sequence. For example, the sequence A-B-C-D has three transitions: A
to B, B to C, and C to D. The graph below shows a subset of the more common
transitions in the crash sequences. It is interpreted as follows: the event listed 
on the x-axis (top) was followed by the event listed on the y-axis. The percentage
shown at the graphical intersection represents the percentage of transitions from one
event (x) to another event (y). For example, 'Fence'
was followed by 'Tree (Standing Only)' in 21% of sequences. Note that we have added a 
state labelled 'Pre-Crash' to help account for sequences with just one event. Another 
notable takeaway is the large number of events that precede Rollover/Overturn.

```{r, fig.height=10, fig.width=12}
my_events %>%
  group_by(year, casenum, veh_no) %>%
  mutate_at("soe", as.character) %>%
  dplyr::rename(event_to = soe) %>%
  mutate(event_from = data.table::shift(event_to, fill = "Pre-Crash")) %>%
  select(event_from, event_to) %>%
  group_by(event_from, event_to) %>% summarize(n=n()) %>%
  group_by(event_from) %>% mutate(n_from = sum(n)) %>%
  mutate(n_pct = n/n_from) %>%
  filter(n_pct>.2, n>5) %>%
  mutate(
    event_from = ifelse(nchar(event_from)>30, paste0(stringr::str_sub(event_from, 1, 30), "..."), event_from),
    #event_to   = paste0(stringr::str_sub(event_to, 1, 30), "..."),
    event_to = stringr::str_wrap(event_to, 40)
    ) %>%
  filter(event_from != event_to) %>%

  ggplot(aes(x=event_from, y=event_to, fill=n_pct, label=scales::percent(n_pct, accuracy = 1))) +
    viridis::scale_fill_viridis() +
    geom_label() +
    scale_x_discrete(position = "top") +
    theme(
      axis.text.x.top = element_text(angle=45, hjust=0),
      axis.ticks = element_blank(),
      #axis.text.x.bottom = element_text(angle=270, hjust = 0, vjust=.5),
      #legend.position = "none"
      )
```
