---
title: "Crash_Sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crash_Sequences}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rfars)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyr)
```

This vignette explores the **Vsoe** (Vehicle Sequence of Events) data to
visualize crash sequence patterns.

**Vsoe** is one of three event-based data files, the others being **Cevent** and
**Vevent**. According to the [data
dictionary](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813254),
**Vevent** "has the same data elements as the **Cevent** data file" plus "a data
element that records the sequential event number for each vehicle," and the
**Vsoe** file "has a subset of the data elements contained in the **Vevent**
data file (it is a simplified **Vevent** data file)" (p. 16). `rfars` therefore
omits **Cevent** and **Vevent**.

## Data

First we download the data. Calling `get_fars(proceed=TRUE)` will download data
from 2015 to 2020 for all states, and store a 'FARS data' folder in your working
directory, with subfolders for raw and prepared files. When `proceed` is set to
`FALSE`, (the default) R will ask your permission before downloading files.

```{r}
get_fars(proceed=TRUE)
```

Once you have the data on your hard drive, you don't need to download it again.
Calling `use_fars()` looks in the working directory for the prepared FARS files,
stacks them, and brings them into the current environment.

```{r}
myFARS <- use_fars()
```

The **Vsoe** data is stored in the `events` tibble of the object returned by
`use_fars()`. Here we see the top 10 individual events:

```{r}
myFARS$events %>%
  group_by(soe) %>% summarize(n=n()) %>%
  arrange(desc(n)) %>%
  slice(1:10) %>%
  
  ggplot(aes(x=n, y=reorder(soe, n), label=scales::comma(n))) +
    geom_col() +
    geom_label()
```

We can also see the top 10 most common sequences:

```{r}
myFARS$events %>%
  select(-aoi) %>%
  pivot_wider(names_from = "veventnum", values_from = "soe", values_fill = "x",
              names_prefix = "event") %>%
  select(starts_with("event")) %>%
  group_by_all() %>%
  summarize(n=n(), .groups = "drop") %>%
  arrange(desc(n)) %>%
  slice(1:10) %>%
  select(event1, event2, n)
  
```

Below we consider all state transitions - the transition from one event to the
next in the sequence. For example, the sequence A-B-C-D has three transitions: A
to B, B to C, and C to D. The graph below shows a subset of the more common
transitions in the crash sequences. It is interpreted as follows: the event on
the y-axis was followed by the event on the x-axis in the percentage of
sequences shown at the intersection of the two events. For example, vehicles
went from 're-entering roadway' to 'cross centerline' in 40% of sequences. Note
that we have added a state labelled 'Pre-Crash' to help account for sequences
with just one event.

```{r, fig.height=7, fig.width=10}
myFARS$events %>%
  group_by(year, state, st_case, veh_no) %>%
  dplyr::rename(event_to = soe) %>%
  mutate(event_from = data.table::shift(event_to, fill = "Pre-Crash")) %>%
  select(event_from, event_to) %>%
  group_by(event_from, event_to) %>% summarize(n=n()) %>%
  group_by(event_from) %>% mutate(n_from = sum(n)) %>%
  mutate(n_pct = n/n_from) %>%
  filter(n_pct>.1, n>500) %>%
  mutate(event_from = stringr::str_wrap(event_from, 40),
         event_to = stringr::str_wrap(event_to, 40)) %>%

  ggplot(aes(y=event_from, x=event_to, fill=n_pct, label=scales::percent(n_pct, accuracy = 1))) +
    viridis::scale_fill_viridis() +
    geom_label() +
    theme(
      axis.text.x.bottom = element_text(angle=270, hjust = 0, vjust=.5),
      legend.position = "none"
      )
```
